ECCCos from the Black Box:\\ Faithful Explanations through\\ Energy-Constrained Conformal Counterfactuals

### 2023-05-11 notebooks/fashion_mnist.qmd
fashion mnist good generative, poor predictive, n1000

# Data:
n_obs = 10000
counterfactual_data = load_fashion_mnist(n_obs)
counterfactual_data.X = pre_process.(counterfactual_data.X)
counterfactual_data.generative_model = vae
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
x_factual = reshape(pre_process(x_factual, noise=0.0f0), input_dim, 1)
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
n_digits = Int(sqrt(input_dim))
output_dim = length(unique(labels))
```

@@

```{julia}
# Model parameters:
epochs = 100
batch_size = minimum([Int(round(n_obs/10)), 128])
n_hidden = 128
activation = Flux.swish
builder = MLJFlux.@builder Flux.Chain(
    Dense(n_in, n_hidden, activation),
    Dense(n_hidden, n_out),
)
n_ens = 5                                   # number of models in ensemble
_loss = Flux.Losses.logitcrossentropy       # loss function
_finaliser = x -> x                         # finaliser function
```

```{julia}
# JEM parameters:
𝒟x = Uniform(-1.0,1.0)
𝒟y = Categorical(ones(output_dim) ./ output_dim)
sampler = ConditionalSampler(
    𝒟x, 𝒟y, 
    input_size=(input_dim,), 
    batch_size=10,
)
α = [1.0,1.0,2e-2]      # penalty strengths
```

@@

# Joint Energy Model:
jem = JointEnergyClassifier(
    sampler;
    builder=builder,
    epochs=epochs,
    batch_size=batch_size,
    finaliser=_finaliser,
    loss=_loss,
    jem_training_params=(
        α=α,verbosity=10,
    ),
    sampling_steps=25,
)

@@

```{julia}
function _plot_eccco_fashion_mnist(
    x::Union{AbstractArray, Int}=x_factual, target::Int=target;
    λ=[0.1,0.25,0.25],
    temp=0.1,
    plt_order = ["MLP", "MLP Ensemble", "JEM", "JEM Ensemble"],
    opt = nothing,
    rng::Union{Int,AbstractRNG}=1234,
    T::Int = 100,
)

@@

```{julia}
λ₁ = 0.1
λ₂ = 0.25
λ₃ = 0.25
Λ = [λ₁, λ₂, λ₃]

>>>>

# Data:
n_obs = 1000
counterfactual_data = load_fashion_mnist(n_obs)
counterfactual_data.X = pre_process.(counterfactual_data.X)
counterfactual_data.generative_model = vae
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
x_factual = reshape(pre_process(x_factual, noise=0.0f0), input_dim, 1)
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
n_digits = Int(sqrt(input_dim))
output_dim = length(unique(labels))
```

@@

```{julia}
# Model parameters:
epochs = 100
batch_size = minimum([Int(round(n_obs/10)), 128])
n_hidden = 512
activation = Flux.swish
builder = MLJFlux.@builder Flux.Chain(
    Dense(n_in, n_hidden, activation),
    Dense(n_hidden, n_hidden, activation),
    Dense(n_hidden, n_hidden, activation),
    Dense(n_hidden, n_out),
)
n_ens = 5                                   # number of models in ensemble
_loss = Flux.Losses.logitcrossentropy       # loss function
_finaliser = x -> x                         # finaliser function
```

```{julia}
# JEM parameters:
𝒟x = Uniform(-1.0,1.0)
𝒟y = Categorical(ones(output_dim) ./ output_dim)
sampler = ConditionalSampler(
    𝒟x, 𝒟y, 
    input_size=(input_dim,), 
    batch_size=10,
)
α = [1.0,1.0,1e-2]      # penalty strengths
```

@@

# Joint Energy Model:
jem = JointEnergyClassifier(
    sampler;
    builder=builder,
    epochs=epochs,
    batch_size=batch_size,
    finaliser=_finaliser,
    loss=_loss,
    jem_training_params=(
        α=α,verbosity=10,
    ),
    sampling_steps=30,
)

@@

```{julia}
function _plot_eccco_fashion_mnist(
    x::Union{AbstractArray, Int}=x_factual, target::Int=target;
    λ=[0.1,0.3,0.3],
    temp=0.1,
    plt_order = ["MLP", "MLP Ensemble", "JEM", "JEM Ensemble"],
    opt = nothing,
    rng::Union{Int,AbstractRNG}=1234,
    T::Int = 100,
)

@@

```{julia}
λ₁ = 0.1
λ₂ = 0.3
λ₃ = 0.3
Λ = [λ₁, λ₂, λ₃]

### 2023-05-11 paper/paper.tex
fashion mnist good generative, poor predictive, n1000

\begin{abstract}
  Counterfactual Explanations offer an intuitive and straightforward way to explain black-box models and offer Algorithmic Recourse to individuals. To address the need for plausible explanations, existing work has primarily relied on surrogate models to learn how the input data is distributed. This effectively reallocates the task of learning realistic representations of the data from the model itself to the surrogate. Consequently, the generated explanations may seem plausible to humans but need not necessarily describe the behaviour of the black-box model faithfully. We formalise this notion of faithfulness through the introduction of a tailored evaluation metric and propose a novel algorithmic framework for generating \textbf{E}nergy-\textbf{C}onstrained \textbf{C}onformal \textbf{Co}unterfactuals that are only as plausible as the model permits. Through extensive empirical studies involving multiple synthetic and real-world datasets, we demonstrate that \textbf{ECCCos} reconcile the need for plausibility and faithfulness. In particular, we show that it is possible to achieve state-of-the-art plausibility without the need for surrogate models. To do so, our framework relies solely on properties defining the black-box model itself by leveraging recent advances in energy-based modelling and conformal inference. While this work is limited in scope and our proposed methodology is only readily applicable to models with gradient access, we anticipate that ECCCo can serve as a baseline for future research directed at providing plausible but faithful Counterfactual Explanations. By highlighting the need for faithfulness in the context of Counterfactual Explanations, we believe that in the short term, our work will enable researchers and practitioners to better distinguish trustworthy from unreliable models. 
\end{abstract}

>>>>

\begin{abstract}
  Counterfactual Explanations offer an intuitive and straightforward way to explain black-box models and offer Algorithmic Recourse to individuals. To address the need for plausible explanations, existing work has primarily relied on surrogate models to learn how the input data is distributed. This effectively reallocates the task of learning realistic explanations for the data from the model itself to the surrogate. Consequently, the generated explanations may seem plausible to humans but need not necessarily describe the behaviour of the black-box model faithfully. We formalise this notion of faithfulness through the introduction of a tailored evaluation metric and propose a novel algorithmic framework for generating \textbf{E}nergy-\textbf{C}onstrained \textbf{C}onformal \textbf{Co}unterfactuals (ECCCos) that are only as plausible as the model permits. Through extensive empirical studies involving multiple synthetic and real-world datasets, we demonstrate that ECCCos reconcile the need for plausibility and faithfulness. In particular, we show that it is possible to achieve state-of-the-art plausibility for models with gradient access without the need for surrogate models. To do so, our framework relies solely on properties defining the black-box model itself by leveraging recent advances in energy-based modelling and conformal prediction. To our knowledge, this is the first venture in this direction for generating faithful Counterfactual Explanations. Thus, we anticipate that ECCCos can serve as a baseline for future research. We believe that our work opens avenues for researchers and practitioners seeking tools to better distinguish trustworthy from unreliable models.
\end{abstract}
